Signed Peripherality Functions for Multivariate Statistical Inference

### 2018-03-30 MS-new/3_Scale.tex
new draft sent to anshu da

\section{Inverse depth and estimation of scatter}
\label{section:dcmSection}
We focus on the estimation of scatter functionals of $\BF$ using peripherality-weighted sign vectors in this section. To this end, in addition to conditions (P) and (A1), we assume the following conditions on the peripherality functions:

@@

Given an inverse depth function $D^-_{\bfX}(\bfx)$ we transform the original random variable as: $\tilde \bfX = D^-_\bfX(\bfx) \bfS(\bfx - \bfmu)$. We attempt to use the transformed data matrix $\tilde \BX = (\tilde \bfx_1, \ldots, \tilde \bfx_n)^T$ as proxy of the original data matrix $\BX$ for improved estimation of the components of the population covariance matrix $\Sigma$ in this section. To this end, in Section~\ref{subsec:dcm} we give some results characterizing the population covariance matrix of these vectors, i.e. $\BV \tilde \bfX$, and its affine equivariant counterpart. {\colrbf details of 3.2 and rest}

@@

To ensure existence and uniqueness of this estimator, we consider the class of scatter estimators $\Sigma_M$ that are obtained as solutions of the following equation:
%
\begin{equation}
E_{\bfZ_M} \left[ u( \| \bfz_M \| )  \frac{\bfz_M \bfz_M^T}{\| \bfz_M \|^2}  - v( \| \bfz_M \| ) \BI_p \right] = 0
\end{equation}
%
with $\bfz_M = \Sigma_M^{-1/2} (\bfx - \bfmu)$. Under the following assumptions on the scalar valued functions $u$ and $v$, the above equation produces a unique solution \citep{HuberBook81}:
%

@@

\subsection{Calculating the sample DCM and ADCM}
Let us denote $\BS(\bfx; \bfmu) = \bfS(\bfx - \bfmu) \bfS(\bfx - \bfmu)^T$. Then, given the depth function and {\it known} location center $\bfmu$, one can show that the vectorized form of $\sqrt n$-times the sample DCM, i.e. $n^{-1/2} \sum_{i=1}^n (D^-_\bfX(\bfx_i))^2 \BS (\bfx_i; \bfmu)$  has an asymptotic multivariate normal distribution with mean $\sqrt n. \ve( \BE [( (D^-_\bfX (\bfX))^2 \BS(\bfx; \bfmu)])$ and some covariance matrix by straightforward application of the central limit theorem (CLT). But in practice the depth at any point $\bfx$ is estimated by the sample depth, i.e. depth calculated using empirical distribution function in place of $\BF$. We also need to replace the known location parameter $\bfmu$ by some estimator $\hat\bfmu_n$.

@@

\subsection{Robust PCA using eigenvectors}

We are mainly interested in using the DCM for robust principal component analysis (PCA). We assume that the eigenvalues of $\Sigma$ are distinct: $\lambda_1 > \lambda_2 > ... > \lambda_p$, for ease of obtaining asymptotic distributions of eigenvectors of the DCM (Appendix \ref{section:appB}). Limiting distributions for the case of non-unique eigenvalues can be obtained following \cite{magyar14}.

@@

\begin{table}[t]
\begin{footnotesize}
    \begin{tabular}{c|cc|ccc|ccc}
    \hline
    4-variate $t_5$    & SCM  & Tyler & HSD-CM & MhD-CM & PD-CM & HSD-wCM & MhD-wCM & PD-wCM \\ \hline
    $n$=20             & 1.04 & 1.02  & 1.10   & 1.07   & 1.02  & 1.09    & 1.07    & 0.98   \\
    $n$=50             & 1.08 & 1.08  & 1.16   & 1.16   & 1.13  & 1.19    & 1.19    & 1.13   \\
    $n$=100            & 1.31 & 1.31  & 1.42   & 1.38   & 1.36  & 1.46    & 1.44    & 1.36   \\
    $n$=300            & 1.46 & 1.54  & 1.81   & 1.76   & 1.95  & 1.88    & 1.88    & 1.95   \\
    $n$=500            & 1.92 & 1.93  & 2.23   & 2.03   & 2.31  & 2.35    & 2.19    & 2.39   \\ \hline
    4-variate $t_6$    & SCM  & Tyler & HSD-CM & MhD-CM & PD-CM & HSD-wCM & MhD-wCM & PD-wCM \\ \hline
    $n$=20             & 1.00 & 1.05  & 1.03   & 1.05   & 1.00  & 1.04    & 1.04    & 0.95   \\
    $n$=50             & 1.03 & 1.01  & 1.13   & 1.12   & 1.11  & 1.19    & 1.17    & 1.10   \\
    $n$=100            & 1.08 & 1.12  & 1.25   & 1.23   & 1.27  & 1.24    & 1.25    & 1.22   \\
    $n$=300            & 1.34 & 1.36  & 1.64   & 1.52   & 1.60  & 1.67    & 1.61    & 1.68   \\
    $n$=500            & 1.26 & 1.34  & 1.55   & 1.49   & 1.60  & 1.65    & 1.61    & 1.69   \\ \hline
    4-variate $t_{10}$ & SCM  & Tyler & HSD-CM & MhD-CM & PD-CM & HSD-wCM & MhD-wCM & PD-wCM \\ \hline
    $n$=20             & 0.90 & 0.89  & 0.95   & 0.98   & 0.98  & 0.96    & 1.01    & 0.95   \\
    $n$=50             & 0.90 & 0.91  & 1.01   & 0.98   & 0.98  & 1.03    & 1.04    & 0.99   \\
    $n$=100            & 0.87 & 0.87  & 0.93   & 0.95   & 1.01  & 0.99    & 1.01    & 1.05   \\
    $n$=300            & 0.87 & 0.87  & 1.09   & 1.09   & 1.17  & 1.14    & 1.16    & 1.23   \\
    $n$=500            & 0.88 & 0.92  & 1.10   & 1.10   & 1.23  & 1.19    & 1.22    & 1.29   \\ \hline
    4-variate $t_{15}$ & SCM  & Tyler & HSD-CM & MhD-CM & PD-CM & HSD-wCM & MhD-wCM & PD-wCM \\ \hline
    $n$=20             & 0.92 & 0.90  & 0.94   & 0.94   & 0.96  & 0.95    & 0.97    & 0.89   \\
    $n$=50             & 0.82 & 0.83  & 0.88   & 0.91   & 0.93  & 0.88    & 0.93    & 0.93   \\
    $n$=100            & 0.84 & 0.87  & 0.92   & 0.95   & 1.00  & 0.93    & 0.96    & 1.00   \\
    $n$=300            & 0.73 & 0.75  & 0.96   & 0.99   & 1.10  & 1.00    & 1.06    & 1.12   \\
    $n$=500            & 0.73 & 0.76  & 0.95   & 0.96   & 1.06  & 0.94    & 0.97    & 1.06   \\ \hline
    4-variate $t_{25}$ & SCM  & Tyler & HSD-CM & MhD-CM & PD-CM & HSD-wCM & MhD-wCM & PD-wCM \\ \hline
    $n$=20             & 0.89 & 0.92  & 0.92   & 0.92   & 0.90  & 0.96    & 0.95    & 0.89   \\
    $n$=50             & 0.82 & 0.84  & 0.89   & 0.90   & 0.91  & 0.93    & 0.96    & 0.92   \\
    $n$=100            & 0.77 & 0.76  & 0.90   & 0.90   & 0.96  & 0.94    & 0.98    & 1.04   \\
    $n$=300            & 0.73 & 0.77  & 0.93   & 0.91   & 0.98  & 1.00    & 0.98    & 1.03   \\
    $n$=500            & 0.67 & 0.71  & 0.83   & 0.83   & 0.96  & 0.88    & 0.90    & 1.00   \\ \hline
    4-variate Normal   & SCM  & Tyler & HSD-CM & MhD-CM & PD-CM & HSD-wCM & MhD-wCM & PD-wCM \\ \hline
    $n$=20             & 0.82 & 0.84  & 0.87   & 0.90   & 0.91  & 0.89    & 0.93    & 0.89   \\
    $n$=50             & 0.80 & 0.81  & 0.87   & 0.88   & 0.88  & 0.88    & 0.92    & 0.88   \\
    $n$=100            & 0.68 & 0.71  & 0.80   & 0.85   & 0.91  & 0.82    & 0.86    & 0.92   \\
    $n$=300            & 0.61 & 0.63  & 0.82   & 0.85   & 0.93  & 0.86    & 0.91    & 0.96   \\
    $n$=500            & 0.60 & 0.64  & 0.77   & 0.80   & 0.90  & 0.82    & 0.86    & 0.96   \\ \hline
    \end{tabular}
\end{footnotesize}
\caption{Finite sample efficiencies of several scatter matrices: $p=4$}
\label{table:FSEtable4}
\end{table}

\subsection{Robust estimation of eigenvalues, and a plug-in estimator of $\Sigma$}

As seen in Theorem~\ref{Theorem:covform}, eigenvalues of the DCM are not same as the population eigenvalues, whereas the ADCM only gives back standardized eigenvalues. However, it is possible to robustly estimate the original eigenvalues by working with the individual columns of the robust score matrix. We do so using the following steps:

@@

{\colrbf (put in 1 or 2 sentences?)}

>>>>

\section{Inverse depth and estimates of scatter}
\label{section:dcmSection}
We focus on the estimation of scatter functionals of $\BF$ using peripherality-weighted sign vectors in this section. To this end, in addition to conditions (P) and (A1), we assume the following conditions on the peripherality functions:

@@

Given an inverse depth function $D^-_{\bfX}(\bfx)$ we transform the original random variable as: $\tilde \bfX = D^-_\bfX(\bfx) \bfS(\bfx - \bfmu)$. We use the transformed data matrix $\tilde \BX = (\tilde \bfx_1, \ldots, \tilde \bfx_n)^T$ as proxy of the original data matrix $\BX$ for improved estimation of the components of the population covariance matrix $\Sigma$ in this section. To this end, in Section~\ref{subsec:dcm} we give some results characterizing the population covariance matrix of these vectors, i.e. $\BV \tilde \bfX$, and its affine equivariant counterpart. Section~\ref{subsec:dcm} gives asymptotic properties of their finite sample versions. In Section~\ref{subsec:robustPCA}, we use eigenvectors of these estimates for robust PCA, deriving their influence functions, as well as asymptotic and finite sample efficiencies. Finally in Section~\ref{subsec:plugin}, we calculate robust estimates of eigenvalues and propose a plug-in robust estimate of $\Sigma$.

@@

To ensure existence and uniqueness of this estimator, we consider the class of scatter estimators $\Sigma_M$ that are obtained as solutions of the following equation:
%
\begin{equation}
\BE_{\bfZ_M} \left[ u( \| \bfz_M \| )  \frac{\bfz_M \bfz_M^T}{\| \bfz_M \|^2}  - v( \| \bfz_M \| ) \BI_p \right] = 0
\end{equation}
%
with $\bfz_M = \Sigma_M^{-1/2} (\bfx - \bfmu)$. Under the following assumptions on the scalar valued functions $u$ and $v$, the above equation produces a unique solution \citep{HuberBook81}:
%

@@

\subsection{Calculating the sample DCM and ADCM}
\label{subsec:sample-DCM}
Let us denote $\BS(\bfx; \bfmu) = \bfS(\bfx - \bfmu) \bfS(\bfx - \bfmu)^T$. Then, given the depth function and {\it known} location center $\bfmu$, one can show that the vectorized form of $\sqrt n$-times the sample DCM, i.e. $n^{-1/2} \sum_{i=1}^n (D^-_\bfX(\bfx_i))^2 \BS (\bfx_i; \bfmu)$  has an asymptotic multivariate normal distribution with mean $\sqrt n. \ve( \BE [( (D^-_\bfX (\bfX))^2 \BS(\bfx; \bfmu)])$ and some covariance matrix by straightforward application of the central limit theorem (CLT). But in practice the depth at any point $\bfx$ is estimated by the sample depth, i.e. depth calculated using empirical distribution function in place of $\BF$. We also need to replace the known location parameter $\bfmu$ by some estimator $\hat\bfmu_n$.

@@

\subsection{Robust PCA using eigenvectors}
\label{subsec:robustPCA}
We are mainly interested in using the DCM for robust principal component analysis (PCA). We assume that the eigenvalues of $\Sigma$ are distinct: $\lambda_1 > \lambda_2 > ... > \lambda_p$, for ease of obtaining asymptotic distributions of eigenvectors of the DCM (Appendix \ref{section:appB}). Limiting distributions for the case of non-unique eigenvalues can be obtained following \cite{magyar14}.

@@

\begin{table}[t]
\begin{scriptsize}
    \begin{tabular}{c|cc|ccc|ccc}
    \hline
    4-variate $t_5$    & SCM  & Tyler & DCM-H & DCM-M & DCM-P & ADCM-H & ADCM-M & ADCM-P \\ \hline
    $n$=20             & 1.04 & 1.02  & 1.10   & 1.07   & 1.02  & 1.09    & 1.07    & 0.98   \\
    $n$=50             & 1.08 & 1.08  & 1.16   & 1.16   & 1.13  & 1.19    & 1.19    & 1.13   \\
    $n$=100            & 1.31 & 1.31  & 1.42   & 1.38   & 1.36  & 1.46    & 1.44    & 1.36   \\
    $n$=300            & 1.46 & 1.54  & 1.81   & 1.76   & 1.95  & 1.88    & 1.88    & 1.95   \\
    $n$=500            & 1.92 & 1.93  & 2.23   & 2.03   & 2.31  & 2.35    & 2.19    & 2.39   \\ \hline
    4-variate $t_6$    & SCM  & Tyler & DCM-H & DCM-M & DCM-P & ADCM-H & ADCM-M & ADCM-P \\ \hline
    $n$=20             & 1.00 & 1.05  & 1.03   & 1.05   & 1.00  & 1.04    & 1.04    & 0.95   \\
    $n$=50             & 1.03 & 1.01  & 1.13   & 1.12   & 1.11  & 1.19    & 1.17    & 1.10   \\
    $n$=100            & 1.08 & 1.12  & 1.25   & 1.23   & 1.27  & 1.24    & 1.25    & 1.22   \\
    $n$=300            & 1.34 & 1.36  & 1.64   & 1.52   & 1.60  & 1.67    & 1.61    & 1.68   \\
    $n$=500            & 1.26 & 1.34  & 1.55   & 1.49   & 1.60  & 1.65    & 1.61    & 1.69   \\ \hline
    4-variate $t_{10}$ & SCM  & Tyler & DCM-H & DCM-M & DCM-P & ADCM-H & ADCM-M & ADCM-P \\ \hline
    $n$=20             & 0.90 & 0.89  & 0.95   & 0.98   & 0.98  & 0.96    & 1.01    & 0.95   \\
    $n$=50             & 0.90 & 0.91  & 1.01   & 0.98   & 0.98  & 1.03    & 1.04    & 0.99   \\
    $n$=100            & 0.87 & 0.87  & 0.93   & 0.95   & 1.01  & 0.99    & 1.01    & 1.05   \\
    $n$=300            & 0.87 & 0.87  & 1.09   & 1.09   & 1.17  & 1.14    & 1.16    & 1.23   \\
    $n$=500            & 0.88 & 0.92  & 1.10   & 1.10   & 1.23  & 1.19    & 1.22    & 1.29   \\ \hline
    4-variate $t_{15}$ & SCM  & Tyler & DCM-H & DCM-M & DCM-P & ADCM-H & ADCM-M & ADCM-P \\ \hline
    $n$=20             & 0.92 & 0.90  & 0.94   & 0.94   & 0.96  & 0.95    & 0.97    & 0.89   \\
    $n$=50             & 0.82 & 0.83  & 0.88   & 0.91   & 0.93  & 0.88    & 0.93    & 0.93   \\
    $n$=100            & 0.84 & 0.87  & 0.92   & 0.95   & 1.00  & 0.93    & 0.96    & 1.00   \\
    $n$=300            & 0.73 & 0.75  & 0.96   & 0.99   & 1.10  & 1.00    & 1.06    & 1.12   \\
    $n$=500            & 0.73 & 0.76  & 0.95   & 0.96   & 1.06  & 0.94    & 0.97    & 1.06   \\ \hline
    4-variate $t_{25}$ & SCM  & Tyler & DCM-H & DCM-M & DCM-P & ADCM-H & ADCM-M & ADCM-P \\ \hline
    $n$=20             & 0.89 & 0.92  & 0.92   & 0.92   & 0.90  & 0.96    & 0.95    & 0.89   \\
    $n$=50             & 0.82 & 0.84  & 0.89   & 0.90   & 0.91  & 0.93    & 0.96    & 0.92   \\
    $n$=100            & 0.77 & 0.76  & 0.90   & 0.90   & 0.96  & 0.94    & 0.98    & 1.04   \\
    $n$=300            & 0.73 & 0.77  & 0.93   & 0.91   & 0.98  & 1.00    & 0.98    & 1.03   \\
    $n$=500            & 0.67 & 0.71  & 0.83   & 0.83   & 0.96  & 0.88    & 0.90    & 1.00   \\ \hline
    4-variate Normal   & SCM  & Tyler & DCM-H & DCM-M & DCM-P & ADCM-H & ADCM-M & ADCM-P \\ \hline
    $n$=20             & 0.82 & 0.84  & 0.87   & 0.90   & 0.91  & 0.89    & 0.93    & 0.89   \\
    $n$=50             & 0.80 & 0.81  & 0.87   & 0.88   & 0.88  & 0.88    & 0.92    & 0.88   \\
    $n$=100            & 0.68 & 0.71  & 0.80   & 0.85   & 0.91  & 0.82    & 0.86    & 0.92   \\
    $n$=300            & 0.61 & 0.63  & 0.82   & 0.85   & 0.93  & 0.86    & 0.91    & 0.96   \\
    $n$=500            & 0.60 & 0.64  & 0.77   & 0.80   & 0.90  & 0.82    & 0.86    & 0.96   \\ \hline
    \end{tabular}
\end{scriptsize}
\caption{Finite sample efficiencies of several scatter matrices: $p=4$. H, M or P after DCM or ADCM indicates the depth function used: H = halfspace depth, M = Mahalanobis depth, P = projection depth.}
\label{table:FSEtable4}
\end{table}

\subsection{Robust estimation of eigenvalues, and a plug-in estimator of $\Sigma$}
\label{subsec:plugin}
As seen in Theorem~\ref{Theorem:covform}, eigenvalues of the DCM are not same as the population eigenvalues, whereas the ADCM only gives back standardized eigenvalues. However, it is possible to robustly estimate the original eigenvalues by working with the individual columns of the robust score matrix. We do so using the following steps:

@@

%{\colrbf (put in 1 or 2 sentences?)}

### 2018-03-30 MS-new/4_SDR.tex
new draft sent to anshu da

\section{Robust PCA and supervised models}

In the presence of a vector of univariate responses, say $\bfy = (y_1,\ldots, y_n)^T$, each $y_i$ being i.i.d. relalizations of a univariate random variable $Y$ that takes values in $\cY \subseteq \BR$, there is substantial literature devoted to utilizing the subspace generated by the basis of $\Sigma = \BV \bfX$ in modelling $\BE(Y|\bfX)$. This ranges from the simple Principal Components Regression (PCR) to Partial Least Squares (PLS) and Envelope methods \citep{Cook10}. In this section, we concentrate on robust inference using Sufficient Dimension Reduction (SDR) \citep{AdragniCook09}, mainly because it provides a general framework for reducing dimensionality of data directly using top eigenvectors of $\Sigma$ (albeit in a different manner than PCR) or an appropriate affine transformation of it.

@@

We formulate a robust version of the above procedure by estimating the quantities $\Gamma_d, \bfmu, \bfmu_y, \sigma^2$ by robust methods. Specifically, we take:
%
\begin{itemize}
\item $\tilde \Gamma_d = $ first $d$ eigenvectors of the sample DCM;
%
\item $\tilde \bfmu = $ spatial median of the rows of $\BX$;
%
\item $\tilde \bfmu_y = $ spatial median of the rows of $\BX_y$, for each $y \in \cY$;
%
\item $\tilde\sigma^2 = \sum_{i=1}^p \tilde s_i/p$, where $\tilde s_i$ is the $i^\text{th}$ diagonal element of $\tilde \BC_y^T \tilde \BC_y/n$, $\tilde \BC_y$ being the matrix composed of rows $\bfx_{yi} - \tilde \bfmu - \tilde \Gamma_d (\tilde \Gamma_d^T \tilde \Gamma_d)^{-1} \tilde \Gamma_d^T (\tilde \bfmu_y - \tilde \bfmu )$ for $i \in \{ 1, \ldots, n \}$ such that $y_i = y$.
\end{itemize}
%
We compare the performance of our robust SDR with the original method of \cite{AdragniCook09} with or without the presence of bad leverage points in $\Sigma$ using a simulation study. For a given value of $p$, we take $n=200, d=1$, generate the responses $y_1, \ldots, y_n$ as independent standard normal, and the predictors as $\bfX_y = {\bf 1} (y + y^2 + y^3) + \bfepsilon$, with ${\bf 1} \in \BR^p$ consisting of all 1's, and $\BV(\bfepsilon) = 25 \BI_p$. We measure performances of the SDR models by their mean squared prediction error on another set of 200 observations $(\bfy^*, \BX^*)$ generated similarly, and taking the average of these errors on 100 such training-test pairs of datasets. Finally we repeat the full setup for different choices of $p = 5,10,25,50,75,100,125,150$.

>>>>

\section{Robust PCA and supervised models}
\label{section:sdrSection}
In the presence of a vector of univariate responses, say $\bfy = (y_1,\ldots, y_n)^T$, each $y_i$ being i.i.d. relalizations of a univariate random variable $Y$ that takes values in $\cY \subseteq \BR$, there is substantial literature devoted to utilizing the subspace generated by the basis of $\Sigma = \BV \bfX$ in modelling $\BE(Y|\bfX)$. This ranges from the simple Principal Components Regression (PCR) to Partial Least Squares (PLS) and Envelope methods \citep{Cook10}. In this section, we concentrate on robust inference using Sufficient Dimension Reduction (SDR) \citep{AdragniCook09}, mainly because it provides a general framework for reducing dimensionality of data directly using top eigenvectors of $\Sigma$ (albeit in a different manner than PCR) or an appropriate affine transformation of it.

@@

We formulate a robust version of the above procedure by estimating the quantities $\Gamma_d, \bfmu, \bfmu_y, \sigma^2$ by robust methods. Specifically, we take:
%
\begin{itemize}
\item $\tilde \Gamma_d = $ first $d$ eigenvectors of the sample DCM;
%
\item $\tilde \bfmu = $ weighted spatial median of the rows of $\BX$;
%
\item $\tilde \bfmu_y = $ weighted spatial median of the rows of $\BX_y$, for each $y \in \cY$;
%
\item $\tilde\sigma^2 = \sum_{i=1}^p \tilde s_i/p$, where $\tilde s_i$ is the $i^\text{th}$ diagonal element of $\tilde \BC_y^T \tilde \BC_y/n$, $\tilde \BC_y$ being the matrix composed of rows $\bfx_{yi} - \tilde \bfmu - \tilde \Gamma_d (\tilde \Gamma_d^T \tilde \Gamma_d)^{-1} \tilde \Gamma_d^T (\tilde \bfmu_y - \tilde \bfmu )$ for $i \in \{ 1, \ldots, n \}$ such that $y_i = y$.
\end{itemize}
%
We compare the performance of our robust SDR with the original method of \cite{AdragniCook09} with or without the presence of bad leverage points in $\Sigma$ using a simulation study. For a given value of $p$, we take $n=200, d=1$, generate the responses $y_1, \ldots, y_n$ as independent standard normal, and the predictors as $\bfX_y = {\bf 1} (y + y^2 + y^3) + \bfepsilon$, with ${\bf 1} \in \BR^p$ consisting of all 1's, and $\BV(\bfepsilon) = 25 \BI_p$. We measure performances of the SDR models by their mean squared prediction error on another set of 200 observations $(\bfy^*, \BX^*)$ generated similarly, and taking the average of these errors on 100 such training-test pairs of datasets. Finally we repeat the full setup for different choices of $p = 5,10,25,50,75,100,125,150$.
